Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-30 Report/Sections/results.tex
Commit before Evan full edit

After fitting the four candidate models to simulated data, we repeat the same exercises and fit all four models to the data shown in Figure \ref{fig:data}. Here we report the results from the full CarHHMM-DFT defined in Section \ref{sec:data} in detail and asses the quality of the fit. We also compare the results of the CarHHMM-DFT with the other candidate models.

The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting, traveling, and searching for prey \citep{Tennessen:2019a}.
Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with behaviours such as foraging and socialization in killer whales \citep{Tennessen:2019a,Tennessen:2019b}. However, \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m, so it is unlikely that any of the 267 dives used in this study are successful foraging dives.

The three subdive behaviours have distinct $Z^{*2}$ values, as the means of each are separated by an order of magnitude. (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean $Z^{*(2)}$ and the smallest variance of $Z^{*(1)}$. It also has the highest auto-correlation in $Z^{*(1)}$. This corresponds to less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive type 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1, and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower in this state. We hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ in subdive state 3 is much higher than the other two states. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between consecutive segments. This corresponds to vigorous swimming activity, especially as the killer whale begins a dive or surfaces from a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix},$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix},$$
%
for the transitions from dive to dive. For dive types 1 and 2,
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for the fine-scale within-dive transitions.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short type 1 dives. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short dives before a killer whale performs a longer, deeper dive. The fine-scale stationary distributions $\hat{\delta^{*(i^*)}}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of 6 selected dives. The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).

@@

Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual of a particular observation is the marginal cumulative distribution function of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}. To easily visualize outliers, this pseudo-residual is often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified. One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution.

We plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}. Observations are weighted by the decoded probability of a particular hidden state, and the histogram is plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}). Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. In addition, $Z^{*(1)}$ has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \cite{Tennessen:2019a}. See the supplementary material for histograms for every observation and hidden state.

@@

The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale. However, it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive level Markov chain or the relationship between the dive level and the subdive level. For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives than when engaged in shorter dives.

@@

The CarHHMM does not use the Fourier sums of the acceleration data, so it fails to decode the subdive states into biologically interpretable results. The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution. See section S-2 for a more complete set of results for each of the candidate models.

>>>>

After fitting the four candidate models to simulated data, we fit the same four models to the data shown in Figure \ref{fig:data}. Here we report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and asses the quality of the fit. We also compare the results of the CarHHMM-DFT with the other candidate models.

The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to these short dives such as resting, traveling, and searching for prey \citep{Tennessen:2019a}.
Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with behaviours such as foraging and socialization in killer whales \citep{Tennessen:2019a,Tennessen:2019b}. However, \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m, so it is unlikely that any of the 267 dives used in this study are successful foraging dives.

The three subdive behaviours have distinct $Z^{*(2)}$ values, as the means of each subdive state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean value of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$. It also has the highest auto-correlation in $Z^{*(1)}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower in this state. We hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between consecutive segments. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions from dive to dive. For dive types 1 and 2,
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for the fine-scale within-dive transitions.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short type 1 dives. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short dives before a killer whale performs a longer, deeper dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of 6 selected dives. The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).

@@

Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}. To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals of this model mostly support that the model is well-specified. One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution.

We also plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}. Observations are weighted by the decoded probability that they correspond to a particular hidden state, and the histogram is finally plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}). Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. Another exception is $Z^{*(1)}$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}. See Section S-2.? for histograms for every observation and hidden state.

@@

The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale. However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive-level Markov chain or the relationship between it and the subdive level. For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.

@@

The CarHHMM does not use the Fourier sums of the acceleration data, so it fails produce biologically interpretable results when decoding the subdive states. The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution. See section S-2 for a more complete set of results for each of the candidate models.

### 2020-07-30 Report/Sections/simulation.tex
Commit before Evan full edit

To compare the performance of the candidate models when the ground-truth is known, we performed a simulation study based on data generated from the full CarHHMM-DFT model. The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}. The ground-truth parameters differ slightly from those in the case study for primarily pedagogical reasons. For example, we set $\Gamma$ such that the simulated data would contain approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states was set to $N^*=2$ to simplify the analysis of the subdive states, and $\mathbf{Z}^{*(2)}$ was changed from a vector to a scalar for additional simplicity.

We fit all four candidate models to the data generated from the CarHHMM-DFT in order to analyze the impact of model mis-specification and compare them using several different metrics. These metrics include the decoding accuracy of the dive and subdive types, bias in parameter estimates, standard error of parameter estimates, and fitting times. In addition, we compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model, which allows us to judge how accurate a given model is when estimating its own uncertainty.

@@

We generated 500 independent data sets from the CarHHMM-DFT. Each data set consisted of a sequence of 100 curves, which we will call a sequence of killer whale dives. Each dive could be one of $N=2$ dive types and the coarse-scale probability transition matrix was set as
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration was gamma distributed, and the coarse-scale emission parameters $\Theta$ were set as
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After the dive durations were calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments, which made up the fine-scale hidden Markov model. The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive were discarded. Each two-second segment was assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
and the observations $Z^*$ were generated using the following parameters:
%
\begin{gather*}
    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
    %
    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
    %
    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
    %
    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
\end{gather*}
%
Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw fine-scale data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one consistent sequence of observations $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.

The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher for subdive type 2. This corresponds to much more vigorous and variable periodic behavior in the acceleration data when the simulated whale is in subdive type 2.

@@

All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) were fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

As expected, the full CarHHMM-DFT was the best-performing model. Given the true subdive type of a particular segment $x^*_{t,t^*} = i$, the CarHHMM-DFT would predict the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations. For the dive types, this average confidence was above 0.9 (see Table \ref{table:accuracy}). Estimates of hidden states were made using the forward-backward algorithm. All parameter estimates had statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased. Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tended to be slight underestimates of the empirical standard error. See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performed almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$. For both subdive types, $\hat \sigma_1^*$ greatly overestimated the true standard deviation. In addition, the empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ were both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3). This implies that parameter estimates can be too confident when auto-correlation is ignored.

The CarHHMM was the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types. The CarHHMM did not have access to the Fourier sums, which were the most informative observations in terms of determining subdive type. It was much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1). The CarHHMM was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy. $\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ were especially poor estimates (see Section S-1.1.3).

Finally, the CarHMM-DFT was comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT was less than half of that of the other models (see Table \ref{table:accuracy}). However, it lacked any hierarchical structure, so it was unable to estimate the coarse-scale dive types. The CarHMM-DFT nonetheless fit a gamma distribution to the dive duration of \textit{all} dives. These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds. See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.

>>>>

To compare the performance of the candidate models when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT model. The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}. The ground-truth parameters differ slightly from those in the case study for primarily pedagogical reasons. For example, we set $\Gamma$ such that the simulated data contains approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ to simplify the analysis of the subdive states, and $Z^{*(2)}$ is changed from a vector to a scalar for additional simplicity.

We fit all four candidate models to the data generated from the CarHHMM-DFT to analyze the impact of model mis-specification and compare them using several different metrics. These metrics include the decoding accuracy of the dive types and subdive states, bias in parameter estimates, standard error of parameter estimates, and fitting times. We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model. This allows us to judge how accurate each model is when estimating its own uncertainty.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive is one of $N=2$ dive types and the coarse-scale probability transition matrix is set as
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are set as
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After the dive durations are calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
    %
    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
    %
    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
    %
    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
\end{gather*}
%
Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one possible realization of $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.

The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2. This corresponds to much more vigorous and variable periodic behavior in the acceleration data.

@@

All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) are fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

As expected, the full CarHHMM-DFT is the best-performing model. Given the true subdive type of a particular segment $X^*_{t,t^*}$, the CarHHMM-DFT predicts the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations. For the dive types, this average confidence was above 0.9 (see Table \ref{table:accuracy}). Estimates of hidden states are made using the forward-backward algorithm. All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased. Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error. See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$. For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation. The empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ are both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3). This implies that parameter estimates can be too confident when auto-correlation is ignored.

The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types. The CarHHMM does not have access to the Fourier sums, which are the most informative observations in terms of determining subdive type. It is much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1). The CarHHMM is also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy. $\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates under the CarHHMM (see Section S-1.1.3).

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, it lacks any hierarchical structure, so it is unable to estimate dive types. The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives. These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds. See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.
